---
title: "Group Project 1"
author: "Huanyu Chen, Shaolei Ma, Ruiqi Xue"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nph)
library(tidyverse)
library(dplyr)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

set.seed(2024)
```

# Proportional-Hazard Assumption

Under proportional-hazards assumption, the hazard function (Cox model) can be written as:
$$h(t|x)=h_0(t)exp(\beta'x)$$
where $t$ is the time, $x$ the vector of covariates, $\beta$ the vector of regression coefficients, $h_0(t)$ is the baseline hazard function. Then, the survival function is
$$S(t|x)=exp[-H_0(t)exp(\beta'x)]$$
where$$H_0(t)=\int_0^th_0(u)du$$
Thus, the distribution function is
$$F(t|x)=1-exp[-H_0(t)exp(\beta'x)]$$
Let $Y$ be a random variable with distribution function $F$, then $U=F(Y)\sim U[0,1],\,(1-U)\sim U[0,1]$, i.e.
$$U=exp[-H_0(t)exp(\beta'x)]\sim U[0,1]$$
if $h_0(t)>0$ for all $t$, then $H_0$ can be inverted and the survival time $T$ of the model can be written as
$$T=H_0^{-1}[-log(U)exp(-\beta'x)]$$
where $U\sim U[0,1]$.

To simply the problem, here we only consider one covariate $x$, which indicates whether the sample belongs to the control arm ($x=0$) or the treatment arm ($x=1$), and set a negative $\beta$ under the assumption that the treatment has a consistent positive effect.

Now, we only need to know $H_0^{-1}$ to simulate the survival time. To do so, we consider two commonly used survival time distributions: exponential and Weibull distribution.

For exponential distribution with scale parameter $\lambda>0$, the possibility density function is $$f_0=\lambda exp(-\lambda t)$$
Then,
$$F_0(t)=1-exp(-\lambda t)$$
$$S_0(t)=1-F_0(t)=exp(-\lambda t)$$
$$H_0(t)=-log(S_0(t))=\lambda t$$
$$h(t)=H_0'(t)=\lambda>0$$
$$H_0^{-1}(t)=\lambda^{-1}t$$
Thus,
$$T=-\lambda^{-1}log(U)exp(-\beta'x)$$
where $U\sim U[0,1]$.

For Weibull distribution with the scale parameter $\lambda$, and is the shape parameter $\gamma$, the possibility density function is $$f_0=\lambda\gamma t^{\gamma-1}exp(-\lambda t^\gamma)$$
Then,
$$F_0(t)=1-exp(-\lambda t^\gamma)$$
$$S_0(t)=1-F_0(t)=exp(-\lambda t^\gamma)$$
$$H_0(t)=-log(S_0(t))=\lambda t^\gamma$$
$$h(t)=H_0'(t)=\lambda\gamma t^{(\gamma-1)}>0$$
$$H_0^{-1}(t)={(\lambda^{-1}t)}^{1/\gamma}$$
Thus,
$$T={(-\lambda^{-1}log(U)exp(-\beta'x))}^{1/\gamma}$$
where $U\sim U[0,1]$.

We can write the simulation process as follows:
```{r}
simulate_func = function(n, baseline, lambda, gamma = NULL, coveff)
{
  # Simulate treatment indicator variable
  x = rbinom(n = n, size = 1, prob = 0.5)
  # Draw from a U(0,1) random variable
  u = runif(n)
  # Simulate survival times depending on the baseline hazard
  if (baseline == "Exponential") {
    t = -log(u) / (lambda * exp(x * coveff))
    # Set the administrative censoring time to guarantee a censor rate of 0.2
    censor_time = qexp(0.8, rate = lambda)
  } else if(baseline == "Weibull") {
    t = (-log(u) / (lambda * exp(x * coveff)))^(1 / gamma)
    censor_time = qweibull(0.8, shape = gamma, scale = lambda)
  }
  # Make event indicator variable applying administrative censoring at t = 5
  d = as.numeric(t < censor_time)
  t = pmin(t, censor_time)
  # Return a tibble object
  if (baseline == "Exponential") {
    return(tibble(x, t, d, n, baseline, lambda, coveff))
  } else if(baseline == "Weibull") {
    return(tibble(x, t, d, n, baseline, lambda, gamma, coveff))
  }
}
```

To observe the potential relevance between test performance and number of samples ($n$), parameter value ($\lambda,\,\gamma$), and coefficient $\beta$, we set $n=50,\,100,\,200$, $\lambda=0,\,0.5,\,1$, $\gamma=0,\,0.5,\,1$, and $\beta=0,\,1,\,2$. We repeat 50 times for each value setting. The generation process is written as follows:

```{r}
exp_param_df = expand.grid(iteration = c(1:50), n = c(50, 100, 200),
            lambda = c(0.5, 0.8, 1), beta = c(-0.5, -1, -5))
wei_param_df = expand.grid(iteration = c(1:50), n = c(50, 100, 200),
            lambda = c(0.5, 0.8, 1), gamma = c(0.5, 0.8, 1), 
            beta = c(-0.5, -1, -5))

exp_results =
  mapply(simulate_func, n = exp_param_df$n, baseline = "Exponential", 
         lambda = exp_param_df$lambda, coveff = exp_param_df$beta)
wei_results = 
  mapply(simulate_func, n = wei_param_df$n, baseline = "Weibull", 
         lambda = wei_param_df$lambda, gamma = wei_param_df$gamma,
         coveff = wei_param_df$beta)

ph_exp_df = tibble()
ph_wei_df = tibble()

for(i in 1:ncol(exp_results))
{
  a = exp_results[, i]
  ph_exp_df = cbind.data.frame(x = a$x, t = a$t, d = a$d, n = a$n,
                               baseline = "Exponential", lambda = a$lambda, 
                               beta = a$coveff) |> as_tibble() |> 
    nest(data = c(x : d)) |> rbind(ph_exp_df)
}

for(i in 1:ncol(wei_results))
{
  a = wei_results[, i]
  ph_wei_df = 
    cbind.data.frame(x = a$x, t = a$t, d = a$d, n = a$n, baseline = "Weibull",
          lambda = a$lambda, gamma = a$gamma, beta = a$coveff) |>
    as_tibble() |> nest(data = c(x : d)) |> rbind(ph_wei_df)
}

ph_exp_df = ph_exp_df |> nest(simulations = c(data))
ph_wei_df = ph_wei_df |> nest(simulations = c(data))
```

Under different settings, we want to test the $H_0:$ there is no difference in survival between the treatment and control arm. Therefore, we use three different log-rank tests and compare the test power at the 0.05 significance level.

```{r}
pwr_func = function(list_df, n = 50)
{
  test1_reject = 0
  test2_reject = 0
  test3_reject = 0
  for(j in 1:nrow(list_df)) {
    dat = list_df |> slice(j) |> unnest(cols = c(data))
    test_results = logrank.maxtest(dat$t, dat$d, dat$x)
    test1_reject = test1_reject + 
      ((test_results$tests |> filter(Test == 1) |> pull(p)) < 0.05)
    test2_reject = test2_reject + 
      ((test_results$tests |> filter(Test == 2) |> pull(p)) < 0.05)
    test3_reject = test3_reject + 
      ((test_results$tests |> filter(Test == 3) |> pull(p)) < 0.05)
  }
  return(
    tibble(
      test1_power = test1_reject / n,
      test2_power = test2_reject / n,
      test3_power = test3_reject / n
    )
  )
}
(ph_exp_df = 
  ph_exp_df |> mutate(power = map(simulations, pwr_func)) |> unnest(power))
(ph_wei_df = 
  ph_wei_df |> mutate(power = map(simulations, pwr_func)) |> unnest(power))
```


(reference link: https://spia.uga.edu/faculty_pages/rbakker/pols8501/OxfordThreeNotes.pdf)

### Cox proportional hazards model
does not assume a particular baseline hazard function $h_0(t)$ but the proportional relation between groups and baseline.
$$
h_i(t) = h_0(t) \exp[\beta_1 X_{i1} + \beta_2 X_{i2} + \ldots + \beta_p X_{ip}]
$$

### Weibull proportional hazards model
assumes a specific functional form for the hazard rate, which can either increase or decrease over time. Its shape parameter distinctly describes whether the hazard rate is increasing, decreasing, or constant.
$$
h_i(t) = \lambda \gamma t^{\gamma - 1} \exp[\beta_1 X_{i1} + \beta_2 X_{i2} + \ldots + \beta_p X_{ip}]
$$
where $\lambda$ is the scale parameter, and $\gamma$ is the shape parameter.


# Non-Proportional-Hazard Assumption
### Weibull accelerated failure time model

# References
Bender, R., Augustin, T., & Blettner, M. (2005). Generating survival times to simulate Cox proportional hazards models. *Statistics in medicine*, 24(11), 1713–1723. https://doi.org/10.1002/sim.2059