---
title: "Group Project 1"
author: "Huanyu Chen, Shaolei Ma, Ruiqi Xue"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nph)
library(tidyverse)
library(dplyr)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Design a set of distributions/models under both proportional-hazard and non-proportional-hazard assumptions
(reference link: https://spia.uga.edu/faculty_pages/rbakker/pols8501/OxfordThreeNotes.pdf)

## Proportional-Hazard Assumption

Under proportional-hazards assumption, the hazard function (Cox model) can be written as:
$$h(t|x)=h_0(t)exp(\beta'x)$$
where $t$ is the time, $x$ the vector of covariates, $\beta$ the vector of regression coefficients, $h_0(t)$ is the baseline hazard function. Then, the survival function is
$$S(t|x)=exp[-H_0(t)exp(\beta'x)]$$
where$$H_0(t)=\int_0^th_0(u)du$$
Thus, the distribution function is
$$F(t|x)=1-exp[-H_0(t)exp(\beta'x)]$$
Let $Y$ be a random variable with distribution function $F$, then $U=F(Y)\sim U[0,1],\,(1-U)\sim U[0,1]$, i.e.
$$U=exp[-H_0(t)exp(\beta'x)]\sim U[0,1]$$
if $h_0(t)>0$ for all $t$, then $H_0$ can be inverted and the survival time $T$ of the model can be written as
$$T=H_0^{-1}[-log(U)exp(-\beta'x)]$$
where $U\sim U[0,1]$.

To simply the problem, here we only consider one covariate $x$, which indicates whether the sample belongs to the control arm ($x=0$) or the treatment arm ($x=1$), and set a negative $\beta$ under the assumption that the treatment has a consistent positive effect.

Now, we only need to know $H_0^{-1}$ to simulate the survival time. To do so, we consider two commonly used survival time distributions: exponential and Weibull distribution.

For exponential distribution with scale parameter $\lambda>0$, the possibility density function is $$f_0=\lambda exp(-\lambda t)$$
Then,
$$F_0(t)=1-exp(-\lambda t)$$
$$S_0(t)=1-F_0(t)=exp(-\lambda t)$$
$$H_0(t)=-log(S_0(t))=\lambda t$$
$$h(t)=H_0'(t)=\lambda>0$$
$$H_0^{-1}(t)=\lambda^{-1}t$$
Thus,
$$T=-\lambda^{-1}log(U)exp(-\beta'x)$$
where $U\sim U[0,1]$.

For Weibull distribution with the scale parameter $\lambda$, and is the shape parameter $\gamma$, the possibility density function is $$f_0=\lambda\gamma t^{\gamma-1}exp(-\lambda t^\gamma)$$
Then,
$$F_0(t)=1-exp(-\lambda t^\gamma)$$
$$S_0(t)=1-F_0(t)=exp(-\lambda t^\gamma)$$
$$H_0(t)=-log(S_0(t))=\lambda t^\gamma$$
$$h(t)=H_0'(t)=\lambda\gamma t^{(\gamma-1)}>0$$
$$H_0^{-1}(t)={(\lambda^{-1}t)}^{1/\gamma}$$
Thus,
$$T={(-\lambda^{-1}log(U)exp(-\beta'x))}^{1/\gamma}$$
where $U\sim U[0,1]$.

We can write the simulation process as follows:
```{r}
simulate_func = function(n, baseline, params = list(), coveff)
{
  # Simulate treatment indicator variable
  x = rbinom(n = n, size = 1, prob = 0.5)
  # Draw from a U(0,1) random variable
  u = runif(n)
  # Simulate survival times depending on the baseline hazard
  if (baseline == "Exponential") {
    t = -log(u) / (params$lambda * exp(x * coveff))
  } else {
    t = (-log(u) / (params$lambda * exp(x * coveff)))^(1 / params$gamma)
  }
  # Make event indicator variable applying administrative censoring at t = 5
  d = as.numeric(t < 5)
  t = pmin(t, 5)
  # Return a data.frame
  data.frame(x = x, t = t, d = d, n = n, baseline = baseline)
}
```

```{r}
dat = simulate_func(n = 100, baseline = "Exponential",
                    params = list(lambda = 0.5), coveff = 0.5)
logrank.test(dat$t, dat$d, dat$x, rho = 0, gamma = 0)
logrank.test(dat$t, dat$d, dat$x, rho = 1, gamma = 0)
logrank.test(dat$t, dat$d, dat$x, rho = 0, gamma = 1)
```


### Cox proportional hazards model
does not assume a particular baseline hazard function $h_0(t)$ but the proportional relation between groups and baseline.
$$
h_i(t) = h_0(t) \exp[\beta_1 X_{i1} + \beta_2 X_{i2} + \ldots + \beta_p X_{ip}]
$$

### Weibull proportional hazards model
assumes a specific functional form for the hazard rate, which can either increase or decrease over time. Its shape parameter distinctly describes whether the hazard rate is increasing, decreasing, or constant.
$$
h_i(t) = \lambda \gamma t^{\gamma - 1} \exp[\beta_1 X_{i1} + \beta_2 X_{i2} + \ldots + \beta_p X_{ip}]
$$
where $\lambda$ is the scale parameter, and $\gamma$ is the shape parameter.


## Non-Proportional-Hazard Assumption
### Weibull accelerated failure time model

# Carry out a simulation study to compare performance of those hypothesis tests in those models.

# References
Bender, R., Augustin, T., & Blettner, M. (2005). Generating survival times to simulate Cox proportional hazards models. *Statistics in medicine*, 24(11), 1713–1723. https://doi.org/10.1002/sim.2059