---
title: "Group Project 1"
author: "Huanyu Chen, Shaolei Ma, Ruiqi Xue"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(nph)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

  options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

set.seed(2024)
```

```{r plot function, include=FALSE}
hazard_plot = function(func1, func2, t_max, title)
{
  ggplot() + 
    xlim(0, t_max) + xlab("t") + ylab("hazard") +
    geom_function(fun = func1, aes(lty = "control")) +
    geom_function(fun = func2, aes(lty = "treatment")) +
    scale_linetype_manual(values = c(1, 2)) +
    labs(title = title)
}
```

# Proportional-Hazard Assumption

Under proportional-hazards assumption, the hazard function (Cox model)
can be written as: $$h(t|x)=h_0(t)exp(\beta'x)$$ where $t$ is the time,
$x$ the vector of covariates, $\beta$ the vector of regression
coefficients, $h_0(t)$ is the baseline hazard function. Then, the
survival function is $$S(t|x)=exp[-H_0(t)exp(\beta'x)]$$
where$$H_0(t)=\int_0^th_0(u)du$$ Thus, the distribution function is
$$F(t|x)=1-exp[-H_0(t)exp(\beta'x)]$$ Let $Y$ be a random variable with
distribution function $F$, then $U=F(Y)\sim U(0,1),\,(1-U)\sim U(0,1)$,
i.e. $$U=exp[-H_0(t)exp(\beta'x)]\sim U(0,1)$$ if $h_0(t)>0$ for all
$t$, then $H_0$ can be inverted and the survival time $T$ of the model
can be written as $$T=H_0^{-1}[-log(U)exp(-\beta'x)]$$ where
$U\sim U(0,1)$.

To simply the problem, here we only consider one covariate $x$, which
indicates whether the sample belongs to the control arm ($x=0$) or the
treatment arm ($x=1$), and set a negative $\beta$ under the assumption
that the treatment has a consistent positive effect.

Now, we only need to know $H_0^{-1}$ to simulate the survival time. To
do so, we consider two commonly used survival time distributions:
exponential and Weibull distribution.

For exponential distribution with scale parameter $\lambda>0$, the
possibility density function is $$f_0=\lambda exp(-\lambda t)$$ Then,
$$F_0(t)=1-exp(-\lambda t)$$ $$S_0(t)=1-F_0(t)=exp(-\lambda t)$$
$$H_0(t)=-log(S_0(t))=\lambda t$$ $$h(t)=H_0'(t)=\lambda>0$$
$$H_0^{-1}(t)=\lambda^{-1}t$$ Thus,
$$T=-\lambda^{-1}log(U)exp(-\beta'x)$$ where $U\sim U(0,1)$.

The hazard function under $\lambda=0.5,\,\beta=-0.5$ could be shown as
follows:

```{r, echo = FALSE}
control_func = function(t) return(0.5)
treatment_func = function(t) return(0.5 * exp(-0.5))

hazard_plot(control_func, treatment_func, 5, 
            "Proportional-Hazard Function for Exponential Models")
```

For Weibull distribution with the scale parameter $\lambda$, and is the
shape parameter $\gamma$, the possibility density function is
$$f_0=\lambda\gamma t^{\gamma-1}exp(-\lambda t^\gamma)$$ Then,
$$F_0(t)=1-exp(-\lambda t^\gamma)$$
$$S_0(t)=1-F_0(t)=exp(-\lambda t^\gamma)$$
$$H_0(t)=-log(S_0(t))=\lambda t^\gamma$$
$$h(t)=H_0'(t)=\lambda\gamma t^{(\gamma-1)}>0$$
$$H_0^{-1}(t)={(\lambda^{-1}t)}^{1/\gamma}$$ Thus,
$$T={(-\lambda^{-1}log(U)exp(-\beta'x))}^{1/\gamma}$$ where
$U\sim U(0,1)$.

The hazard function under $\lambda=0.5,\,\gamma=1.2,\,\beta=-0.5$ could
be shown as follows:

```{r, echo = FALSE}
control_func = function(t) return(0.5 * 1.2 * t ^ (1.2 - 1))
treatment_func = function(t)
  return(0.5 * 1.2 * t ^ (1.2 - 1) * exp(-0.5))

hazard_plot(control_func, treatment_func, 5,
            "Proportional-Hazard Function for Weibull Models")
```

We can write the simulation process as follows:

```{r}
ph_simulate_func = function(n, baseline, lambda, gamma = NULL, coveff)
{
  # Simulate treatment indicator variable
  x = rbinom(n = n, size = 1, prob = 0.5)
  # Draw from a U(0,1) random variable
  u = runif(n)
  # Simulate survival times depending on the baseline hazard
  if (baseline == "Exponential") {
    t = -log(u) / (lambda * exp(x * coveff))
    # Set the administrative censoring time to guarantee a censor rate of 0.2
    censor_time = qexp(0.8, rate = lambda)
  } else if(baseline == "Weibull") {
    t = (-log(u) / (lambda * exp(x * coveff)))^(1 / gamma)
    censor_time = qweibull(0.8, shape = gamma, scale = 1 / lambda)
  }
  # Make event indicator variable applying administrative censoring
  d = as.numeric(t < censor_time)
  t = pmin(t, censor_time)
  # Return a tibble object
  if (baseline == "Exponential") {
    return(tibble(x, t, d, n, baseline, lambda, coveff))
  } else if(baseline == "Weibull") {
    return(tibble(x, t, d, n, baseline, lambda, gamma, coveff))
  }
}
```

To observe the potential relevance between test performance and number
of samples ($n$), parameter value ($\lambda,\,\gamma$), and coefficient
$\beta$, we set $n=100,\,200$, $\lambda=0.5,0.8,1$, $\gamma=1.2,\,1.5$,
and $\beta=-5,-1,-0.5$. We repeat 50 times for each value setting. The
generation process is written as follows:

```{r}
exp_param_df = expand.grid(iteration = c(1:50), n = c(100, 200),
            lambda = c(0.5, 0.8, 1), beta = c(-0.5, -1, -5))
wei_param_df = expand.grid(iteration = c(1:50), n = c(100, 200),
            lambda = c(0.5, 1), gamma = c(1.2, 1.5), 
            beta = c(-1, -5))

exp_results =
  mapply(ph_simulate_func, n = exp_param_df$n, baseline = "Exponential", 
         lambda = exp_param_df$lambda, coveff = exp_param_df$beta)
wei_results = 
  mapply(ph_simulate_func, n = wei_param_df$n, baseline = "Weibull", 
         lambda = wei_param_df$lambda, gamma = wei_param_df$gamma,
         coveff = wei_param_df$beta)

ph_exp_df = tibble()
ph_wei_df = tibble()

for(i in 1:ncol(exp_results))
{
  a = exp_results[, i]
  ph_exp_df = cbind.data.frame(x = a$x, t = a$t, d = a$d, n = a$n,
                               baseline = "Exponential", lambda = a$lambda, 
                               beta = a$coveff) |> as_tibble() |> 
    nest(data = c(x : d)) |> rbind(ph_exp_df)
}

for(i in 1:ncol(wei_results))
{
  a = wei_results[, i]
  ph_wei_df = 
    cbind.data.frame(x = a$x, t = a$t, d = a$d, n = a$n, baseline = "Weibull",
          lambda = a$lambda, gamma = a$gamma, beta = a$coveff) |>
    as_tibble() |> nest(data = c(x : d)) |> rbind(ph_wei_df)
}

ph_exp_df = ph_exp_df |> nest(simulations = c(data))
ph_wei_df = ph_wei_df |> nest(simulations = c(data))
```

Under different settings, we want to test the $H_0:$ there is no
difference in survival between the treatment and control arm. Therefore,
we use three different log-rank tests and compare the test power at the
0.05 significance level.

```{r}
specif_func = function(list_df, n = 50)
{
  test1_reject = 0
  test2_reject = 0
  test3_reject = 0
  for(j in 1:nrow(list_df)) {
    dat = list_df |> slice(j) |> unnest(cols = c(data))
    test_results = logrank.maxtest(dat$t, dat$d, dat$x)
    test1_reject = test1_reject + 
      ((test_results$tests |> filter(Test == 1) |> pull(p)) < 0.05)
    test2_reject = test2_reject + 
      ((test_results$tests |> filter(Test == 2) |> pull(p)) < 0.05)
    test3_reject = test3_reject + 
      ((test_results$tests |> filter(Test == 3) |> pull(p)) < 0.05)
  }
  return(
    tibble(
      test1_specificity = test1_reject / n,
      test2_specificity = test2_reject / n,
      test3_specificity = test3_reject / n
    )
  )
}


ph_exp_df = ph_exp_df |>
  mutate(specificity = map(simulations, specif_func)) |>
  unnest(specificity) |>
  select(-simulations)


ph_wei_df = ph_wei_df |> 
  mutate(specificity = map(simulations, specif_func)) |> 
  unnest(specificity) |>
  select(-simulations)

```

\newpage

```{r}
kable(ph_exp_df, caption = "Specificity of 3 Log-Rank Tests based on PH Assumption")
```

```{r}
kable(ph_wei_df, caption = "Specificity of 3 Log-Rank Tests based on PH Assumption")
```

```{r}
ph_exp_iddf = ph_exp_df |>
  mutate(id = seq_len(nrow(ph_exp_df ))) |>
  select(id, everything()) |>
  pivot_longer(cols = starts_with("test"), names_to = "Test", values_to = "Specificity") |>
  mutate(Test = str_remove(Test, "_specificity"))

ph_exp_power_graph = ggplot(ph_exp_iddf, aes(x = id, y = Specificity, color = Test)) +
  geom_line() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", alpha = 0.5) +  
  annotate("text", x = 1.3, y = 1, label = "beta = -5.0", vjust = -0.2, hjust = 0, alpha = 0.7) +
  geom_vline(xintercept = 7, linetype = "dashed", color = "black", alpha = 0.5) +  
  annotate("text", x = 7.3, y = 1, label = "beta = -1.0", vjust = -0.2, hjust = 0, alpha = 0.7) +
  geom_vline(xintercept = 13, linetype = "dashed", color = "black", alpha = 0.5) +  
  annotate("text", x = 13.3, y = 1, label = "beta = -0.5", vjust = -0.2, hjust = 0, alpha = 0.7) +
  labs(
    x = "id",
    y = "Specificity",
    title = "Specificity of Three Tests over combinations of parameters",
    color = "Test"
  )

ph_exp_power_graph


ph_wei_iddf = ph_wei_df |>
  mutate(id = seq_len(nrow(ph_wei_df ))) |>
  select(id, everything()) |>
  pivot_longer(cols = starts_with("test"), names_to = "Test", values_to = "Specificity") |>
  mutate(Test = str_remove(Test, "_specificity"))

ph_wei_spe_graph = ggplot(ph_wei_iddf, aes(x = id, y = Specificity, color = Test)) +
  geom_line() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", alpha = 0.5) +
  annotate("text", x = 1.3, y = 1, label = "beta = -5.0", vjust = -0.2, hjust = 0, alpha = 0.7) +
  geom_vline(xintercept = 13, linetype = "dashed", color = "black", alpha = 0.5) +
  annotate("text", x = 13.3, y = 1, label = "beta = -1.0", vjust = -0.2, hjust = 0, alpha = 0.7) +
  geom_vline(xintercept = 25, linetype = "dashed", color = "black", alpha = 0.5) +
  annotate("text", x = 25.3, y = 1, label = "beta = -0.5", vjust = -0.2, hjust = 0, alpha = 0.7) +
  labs(
    x = "id",
    y = "Specificity",
    title = "Specificity of Three Tests over combinations of parameters",
    color = "Test"
  )

ph_wei_spe_graph

```

# Non-Proportional-Hazard Assumption

Under Non-Proportional-Hazard Assumption, we still consider the
exponential model and Weibull model.

## Piecewise Exponential Model

To simplify the problem, we set the baseline hazard function to be a
constant $\lambda_0=0.5$, which indicates that the survival time for the
control arm follows exponential distribution.

### Late Effect

For the treatment arm, we suppose the hazard function for the treatment
arm is: $$h(t|x=1)=\left\{
\begin{aligned}
\lambda_0 & & t<1 \\ \lambda_1 & & t\geq 1
\end{aligned}
\right.$$ Then, $$H(t|x=1)=\left\{
\begin{aligned}
\lambda_0 t & & t<1 \\ (\lambda_0+\lambda_1)t-\lambda_1 & & t\geq 1
\end{aligned}
\right.$$ $$S(t|x=1)=exp(-H(t|x=1))=\left\{
\begin{aligned}
exp(-\lambda_0 t) & & t<1 \\ exp(-(\lambda_0+\lambda_1)t+\lambda_1) & & t\geq 1
\end{aligned}
\right.$$ $$F(t|x=1)=1-S(t|x=1)=\left\{
\begin{aligned}
1-exp(-\lambda_0 t) & & t<1 \\ 1-exp(-(\lambda_0+\lambda_1)t+\lambda_1) & & t\geq 1
\end{aligned}
\right.$$ Let $1-U=F(t|x=1)$, then $(1-U)\sim U(0,1)$,
$U=S(t|x=1)\sim U(0,1)$. Thus, $$T=\left\{
\begin{aligned}
-\lambda_0^{-1}log(U) & & U>exp(-\lambda_0) \\
\frac{\lambda_1-log(U)}{\lambda_0+\lambda_1} & & U\leq exp(-\lambda_0)
\end{aligned}
\right.$$

The hazard function under $\lambda_0=0.5,\,\lambda_1=0.3$ could be shown
as follows:

```{r, echo = FALSE}
control_func = function(t) return(0.5)
treatment_func = function(t) return(0.5 * (t < 1) + 0.3 * (t >= 1))

hazard_plot(control_func, treatment_func, qexp(0.8, rate = 0.5),
            "Non-Proportional-Hazard Function for Late-Effect Exponential Models")
```

With the distribution function of survival times, we can write the
simulation process as follows (note: for early effect piecewise models,
the expression for all functions are similar except for the definition
domains, so the simulation process is similar and we write it down as
well.)

```{r}
piecewise_sim_func = function(n, lambda0 = 0.5, lambda1, type)
{
  # Set the administrative censoring time to guarantee a censor rate of 0.2 for control arm
  censor_time = qexp(0.8, rate = lambda0)
  
  u0 = runif(n)
  t0 = - log(u0) / lambda0
  u1 = runif(n)
  if(type == "late")
    t1 = (u1 > exp(-lambda0)) * (-log(u1) / lambda0) +
    (u1 <= exp(-lambda0)) * ((lambda1 - log(u1)) / (lambda0 + lambda1))
  else if(type == "early")
    t1 = (u1 <= exp(-lambda0)) * (-log(u1) / lambda0) +
    (u1 > exp(-lambda0)) * ((lambda1 - log(u1)) / (lambda0 + lambda1))
  
  # Make event indicator variable applying administrative censoring
  d0 = as.numeric(t0 < censor_time)
  d1 = as.numeric(t1 < censor_time)
  t0 = pmin(t0, censor_time)
  t1 = pmin(t1, censor_time)
  
  control_df = tibble(x = rep(0, n), t = t0, d = d0, n, lambda0, lambda1)
  treat_df = tibble(x = rep(1, n), t = t1, d = d1, n, lambda0, lambda1)
  return(rbind(control_df, treat_df))
}
```

```{r}
late_pw_param_df = expand.grid(iteration = c(1:50), n = c(100, 200),
            lambda0 = c(0.5, 0.8), lambda1 = c(0.3, 0.4))

late_pw_results =
  mapply(piecewise_sim_func, n = late_pw_param_df$n,
         lambda0 = late_pw_param_df$lambda0, lambda1 = late_pw_param_df$lambda1,
         type = "late")

late_pw_df = tibble()

for(i in 1:ncol(late_pw_results))
{
  a = late_pw_results[, i]
  late_pw_df = cbind.data.frame(x = a$x, t = a$t, d = a$d, n = a$n, 
                                lambda0 = a$lambda0, lambda1 = a$lambda1) |> 
    as_tibble() |> nest(data = c(x : d)) |> rbind(late_pw_df)
}

late_pw_df = late_pw_df |> nest(simulations = c(data))
```

Under different settings, we want to test the $H_0:$ there is no
difference in survival between the treatment and control arm. Therefore,
we use three different log-rank tests and compare the test power at the
0.05 significance level.

```{r}
late_pw_df = late_pw_df |>
  mutate(specificity = map(simulations, specif_func)) |>
  unnest(specificity) |>
  select(-simulations)

```

```{r}

kable(late_pw_df, caption = "Specificity of 3 Log-Rank Tests based on NPH Assumption (Late)")
```

```{r}
late_pw_iddf = late_pw_df |>
  mutate(id = seq_len(nrow(late_pw_df))) |>
  select(id, everything()) |>
  pivot_longer(cols = starts_with("test"), names_to = "Test", values_to = "Specificity") |>
  mutate(Test = str_remove(Test, "_specificity"))

late_pw_spe_graph = ggplot(late_pw_iddf, aes(x = id, y = Specificity, color = Test)) +
  geom_line() +
  geom_vline(xintercept = 3, linetype = "dashed", color = "black", alpha = 0.5) +
  annotate("text", x = 3.05, y = 1, label = "n=200, lambda0=0.5,\nlambda1=0.4", vjust = 0.5, hjust = 0, alpha = 0.7) +
  geom_vline(xintercept = 7, linetype = "dashed", color = "black", alpha = 0.5) +
  annotate("text", x = 7.05, y = 1, label = "n=200, lambda0=0.5,\nlambda1=0.4", vjust = 0.5, hjust = 0, alpha = 0.7) +
  labs(
    x = "id",
    y = "Specificity",
    title = "Specificity of Three Tests over combinations of parameters",
    color = "Test"
  )

late_pw_spe_graph
```

### Early Effect

We can use the similar simulation method to generate piecewise
exponential models in which the treatment arm shows early effect. The
hazard function becomes: $$h(t|x=1)=\left\{
\begin{aligned}
\lambda_0 & & t\geq 1 \\ \lambda_1 & & t<1
\end{aligned}
\right.$$ Similarly, it can be derived that $$T=\left\{
\begin{aligned}
-\lambda_0^{-1}log(U) & & U\leq exp(-\lambda_0) \\
\frac{\lambda_1-log(U)}{\lambda_0+\lambda_1} & & U>exp(-\lambda_0)
\end{aligned}
\right.$$ where $U\sim U(0,1)$.

The hazard function under $\lambda_0=0.5,\,\lambda_1=0.3$ could be shown
as follows:

```{r, echo = FALSE}
control_func = function(t) return(0.5)
treatment_func = function(t) return(0.5 * (t >= 1) + 0.3 * (t < 1))

hazard_plot(control_func, treatment_func, qexp(0.8, rate = 0.5),
            "Non-Proportional-Hazard Function for Early-Effect Exponential Models")
```

```{r}
early_pw_param_df = expand.grid(iteration = c(1:50), n = c(100, 200),
            lambda0 = c(0.8, 0.9), lambda1 = c(0.6, 0.7))

early_pw_results =
  mapply(piecewise_sim_func, n = early_pw_param_df$n,
         lambda0 = early_pw_param_df$lambda0, lambda1 = early_pw_param_df$lambda1,
         type = "early")

early_pw_df = tibble()

for(i in 1:ncol(early_pw_results))
{
  a = early_pw_results[, i]
  early_pw_df = cbind.data.frame(x = a$x, t = a$t, d = a$d, n = a$n, 
                                lambda0 = a$lambda0, lambda1 = a$lambda1) |> 
    as_tibble() |> nest(data = c(x : d)) |> rbind(early_pw_df)
}

early_pw_df = early_pw_df |> nest(simulations = c(data))
```

Under different settings, we want to test the $H_0:$ there is no
difference in survival between the treatment and control arm. Therefore,
we use three different log-rank tests and compare the test power at the
0.05 significance level.

```{r}
early_pw_df = early_pw_df |> 
  mutate(specificity = map(simulations, specif_func)) |> 
  unnest(specificity) |>
  select(-simulations)
```

```{r}
kable(early_pw_df, caption = "Specificity of 3 Log-Rank Tests based on NPH Assumption (Early)")
```

```{r}
early_pw_iddf = early_pw_df |>
  mutate(id = seq_len(nrow(early_pw_df))) |>
  select(id, everything()) |>
  pivot_longer(cols = starts_with("test"), names_to = "Test", values_to = "Specificity") |>
  mutate(Test = str_remove(Test, "_specificity"))

early_pw_spe_graph = ggplot(early_pw_iddf, aes(x = id, y = Specificity, color = Test)) +
  geom_line() +
  geom_vline(xintercept = 3, linetype = "dashed", color = "black", alpha = 0.5) +
  annotate("text", x = 3.05, y = 1, label = "n=200, lambda0=0.8,\nlambda1=0.7", vjust = 0.7, hjust = 0, alpha = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", alpha = 0.5) +
  annotate("text", x = 1.05, y = 1, label = "n=200,\nlambda0=0.9,\nlambda1=0.7", vjust = 0.8, hjust = 0, alpha = 0.7) +
  labs(
    x = "id",
    y = "Specificity",
    title = "Specificity of Three Tests over combinations of parameters",
    color = "Test"
  )

early_pw_spe_graph
```

## Weibull Model

To simplify the problem, we assume the control and treatment arm share
the same scale parameter $\lambda$. For the control arm, suppose the
hazard function is: $$h(t|x=0)=\lambda\gamma_0 t^{(\gamma_0-1)}.$$ Then,
$$H(t|x=0)=\lambda t^\gamma_0$$
$$S(t|x=0)=exp(-H(t|x=0))=exp(-\lambda t^\gamma_0)$$
$$F(t|x=0)=1-S(t|x=0)=1-exp(-\lambda t^\gamma_0)$$ Let $1-U=F(t|x=0)$,
then $(1-U)\sim U(0,1)$, $U=S(t|x=0)\sim U(0,1)$. Thus,
$$T={(-\lambda^{-1}log(U))}^{1/\gamma_0}$$

Similarly, we can write the hazard function for the treatment arm as:
$$h(t|x=1)=\lambda\gamma_1 t^{(\gamma_1-1)}.$$ We can derive that
$$T={(-\lambda^{-1}log(U))}^{1/\gamma_1}$$

### Late Effect

The hazard function under $\lambda=0.2,\,\gamma_0=3,\,\gamma_1=2.5$
could be shown as follows:

```{r, echo = FALSE}
control_func = function(t) return(0.2 * 3 * t ^ (3 - 1))
treatment_func = function(t) return(0.2 * 2.5 * t ^ (2.5 - 1))

hazard_plot(control_func, treatment_func, 
            qweibull(0.8, shape = 3, scale = 1 / 0.2),
            "Non-Proportional-Hazard Function for Late-Effect Weibull Models")
```

With the distribution function of survival times, we can write the
simulation process as follows.

```{r}
weibull_sim_func = function(n, lambda = 0.5, gamma0, gamma1)
{
  # Set the administrative censoring time to guarantee a censor rate of 0.2 for control arm
  censor_time = qweibull(0.8, shape = gamma0, scale = 1 / lambda)
  
  u0 = runif(n)
  t0 = (- log(u0) / lambda) ^ (1 / gamma0)
  u1 = runif(n)
  t1 = (- log(u1) / lambda) ^ (1 / gamma1)
  
  # Make event indicator variable applying administrative censoring
  d0 = as.numeric(t0 < censor_time)
  d1 = as.numeric(t1 < censor_time)
  t0 = pmin(t0, censor_time)
  t1 = pmin(t1, censor_time)
  
  control_df = tibble(x = rep(0, n), t = t0, d = d0, 
                      n, lambda, gamma0, gamma1)
  treat_df = tibble(x = rep(1, n), t = t1, d = d1, 
                    n, lambda, gamma0, gamma1)
  return(rbind(control_df, treat_df))
}
```

```{r}
late_wei_param_df = 
  expand.grid(iteration = c(1:50), n = c(100, 200), 
              lambda = c(0.2, 0.4), gamma0 = c(4, 5), 
              gamma1 = c(2.5, 3))

late_wei_results =
  mapply(weibull_sim_func, n = late_wei_param_df$n,
         lambda = late_wei_param_df$lambda,
         gamma0 = late_wei_param_df$gamma0, gamma1 = late_wei_param_df$gamma1)

late_wei_df = tibble()

for(i in 1:ncol(late_wei_results))
{
  a = late_wei_results[, i]
  late_wei_df = cbind.data.frame(x = a$x, t = a$t, d = a$d, n = a$n, 
                                lambda = a$lambda, gamma0 = a$gamma0,
                                gamma1 = a$gamma1) |> 
    as_tibble() |> nest(data = c(x : d)) |> rbind(late_wei_df)
}

late_wei_df = late_wei_df |> nest(simulations = c(data))
```

Under different settings, we want to test the $H_0:$ there is no
difference in survival between the treatment and control arm. Therefore,
we use three different log-rank tests and compare the test power at the
0.05 significance level.

```{r}
late_wei_df = late_wei_df |>
  mutate(specificity = map(simulations, specif_func)) |>
  unnest(specificity) |>
  select(-simulations)
```

```{r}
kable(late_wei_df, caption = "Specificity of 3 Log-Rank Tests based on NPH Assumption")
```

```{r}
late_wei_iddf = late_wei_df |>
  mutate(id = seq_len(nrow(late_wei_df))) |>
  select(id, everything()) |>
  pivot_longer(cols = starts_with("test"), names_to = "Test", values_to = "Specificity") |>
  mutate(Test = str_remove(Test, "_specificity"))

late_wei_spe_graph = ggplot(late_wei_iddf, aes(x = id, y = Specificity, color = Test)) +
  geom_line() +
  #geom_smooth(se=FALSE) +
  labs(
    x = "id",
    y = "Specificity",
    title = "Specificity of Three Tests over combinations of parameters",
    color = "Test"
  )
# late_wei_spe_graph = late_wei_spe_graph +
#   geom_point(data = subset(late_wei_iddf, Specificity > 0.9), aes(color = Test), size = 2)

# late_wei_spe_graph = late_wei_spe_graph +
#   geom_point(data = subset(late_wei_iddf, Specificity < 0.2), aes(color = Test), size = 1.5)

# late_wei_spe_graph = late_wei_spe_graph + 
#   geom_vline(xintercept = 3, linetype = "dashed", color = "black", alpha = 0.5) +
#   annotate("text", x = 3.1, y = 1, label = "n=200, lambda=0.5,\ngamma0=5, gamma1=1.5", vjust = 0.2, hjust = 0, alpha = 0.7) +
#   geom_vline(xintercept = 1, linetype = "dashed", color = "black", alpha = 0.5) +
#   annotate("text", x = 1.1, y = 1, label = "n=200,\nlambda=0.8,\ngamma0=5,\ngamma1=1.5", vjust = 2, hjust = 0, alpha = 0.7) 

late_wei_spe_graph
```

### Early Effect

The hazard function under $\lambda=0.5,\,\gamma_0=2,\,\gamma_1=1.5$
could be shown as follows:

```{r, echo = FALSE}
control_func = function(t) return(0.7)
treatment_func = function(t) return(0.5 * 1.2 * t ^ (1.2 - 1))

hazard_plot(control_func, treatment_func, 
            qexp(0.8, rate = 0.7),
            "Non-Proportional-Hazard Function for Early-Effect Weibull Models")
```

With the distribution function of survival times, we can write the
simulation process as follows.

```{r}
weiexp_sim_func = function(n, lambda0, lambda1 = 0.5, gamma)
{
  # Set the administrative censoring time to guarantee a censor rate of 0.2 for control arm
  censor_time = qexp(0.8, rate = lambda0)
  
  u0 = runif(n)
  t0 = - log(u0) / lambda0
  u1 = runif(n)
  t1 = (- log(u1) / lambda1) ^ (1 / gamma)
  
  # Make event indicator variable applying administrative censoring
  d0 = as.numeric(t0 < censor_time)
  d1 = as.numeric(t1 < censor_time)
  t0 = pmin(t0, censor_time)
  t1 = pmin(t1, censor_time)
  
  control_df = tibble(x = rep(0, n), t = t0, d = d0, 
                      n, lambda0, lambda1, gamma)
  treat_df = tibble(x = rep(1, n), t = t1, d = d1, 
                    n, lambda0, lambda1, gamma)
  return(rbind(control_df, treat_df))
}
```

```{r}
early_wei_param_df = 
  expand.grid(iteration = c(1:50), n = c(100, 200), 
              lambda0 = c(0.8, 1), lambda1 = 0.5, gamma = c(1.1, 1.2))

early_wei_results =
  mapply(weiexp_sim_func, n = early_wei_param_df$n,
         lambda0 = early_wei_param_df$lambda0,
         lambda1 = early_wei_param_df$lambda1, gamma = early_wei_param_df$gamma)

early_wei_df = tibble()

for(i in 1:ncol(early_wei_results))
{
  a = early_wei_results[, i]
  early_wei_df = cbind.data.frame(x = a$x, t = a$t, d = a$d, n = a$n, 
                                lambda0 = a$lambda0, lambda1 = a$lambda1,
                                gamma = a$gamma) |> 
    as_tibble() |> nest(data = c(x : d)) |> rbind(early_wei_df)
}

early_wei_df = early_wei_df |> nest(simulations = c(data))
```

Under different settings, we want to test the $H_0:$ there is no
difference in survival between the treatment and control arm. Therefore,
we use three different log-rank tests and compare the test power at the
0.05 significance level.

```{r}
early_wei_df = early_wei_df |>
  mutate(specificity = map(simulations, specif_func)) |>
  unnest(specificity) |>
  select(-simulations)
```

```{r}
kable(early_wei_df, caption = "Specificity of 3 Log-Rank Tests based on NPH Assumption")
```

```{r}
early_wei_iddf = early_wei_df |>
  mutate(id = seq_len(nrow(early_wei_df))) |>
  select(id, everything()) |>
  pivot_longer(cols = starts_with("test"), names_to = "Test", values_to = "Specificity") |>
  mutate(Test = str_remove(Test, "_specificity"))

early_wei_spe_graph = ggplot(early_wei_iddf, aes(x = id, y = Specificity, color = Test)) +
  geom_line() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", alpha = 0.5) +
  annotate("text", x = 1.05, y = 1, label = "n=200,\nlambda0=1.0,\nlambda1=0.5,\ngamma=1.2", vjust = 1, hjust = 0, alpha = 0.7) +
  geom_vline(xintercept = 4, linetype = "dashed", color = "black", alpha = 0.5) +
  annotate("text", x = 4.05, y = 1, label = "n=100,\nlambda0=0.8,\nlambda1=0.5,\ngamma=1.2", vjust = 1, hjust = 0, alpha = 0.7) +
  labs(
    x = "id",
    y = "Specificity",
    title = "Specificity of Three Tests over combinations of parameters",
    color = "Test"
  )
early_wei_spe_graph
```

```{r}
ph_exp_power_graph
ph_wei_spe_graph
early_pw_spe_graph
late_pw_spe_graph

par(mfrow=c(2,2))
```

