---
title: "Group Project 1"
author: "Huanyu Chen, Shaolei Ma, Ruiqi Xue"
date: "2024-02-20"
output:
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
library(nph)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(gridExtra)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = .3,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

set.seed(2024)
```

```{r plot function, include=FALSE}
hazard_plot = function(func1, func2, t_max, title)
{
  ggplot() + 
    xlim(0, t_max) + xlab("t") + ylab("hazard") +
    geom_function(fun = func1, aes(lty = "control")) +
    geom_function(fun = func2, aes(lty = "treatment")) +
    scale_linetype_manual(values = c(1, 2)) +
    labs(title = title)
}
```

```{=tex}
\begin{abstract}
This study compares the performance of three hypothesis tests for time-to-event data: the conventional log-rank test and two variants of the weighted log-rank test. We assess their effectiveness under scenarios involving both proportional and non-proportional hazard functions, using Monte Carlo simulation techniques to evaluate power across a range of coefficient values. Our analysis highlights nuanced differences in their ability to detect treatment effects, providing insights into selecting appropriate statistical methodologies for analyzing time-to-event data in clinical trials.
\end{abstract}
```
# Introduction

In clinical trials, treatment efficacy is typically evaluated using time-to-event outcomes and the hazard ratio under the proportional hazards assumption. The log-rank test is used to compare observed and expected event counts. However, in real-world scenarios, the proportional hazards assumption may not always hold, requiring statistical adjustments. To address this issue, researchers have proposed weighted log-rank tests that incorporate parameters for different emphases on early or late events. To test the performance of various log-rank tests, we focus on both proportional hazards and non-proportional hazards based on exponential and Weibull distribution assumptions, and then evaluate the power of different parameters.

# Methods

## Proportional-Hazard Assumption

Under proportional-hazards assumption, the hazard function (Cox model) can be written as: $$h(t|x)=h_0(t)exp(\beta'x)$$ where $t$ is the time, $x$ the vector of covariates, $\beta$ the vector of regression coefficients, $h_0(t)$ is the baseline hazard function. Then, the survival function is $$S(t|x)=exp[-H_0(t)exp(\beta'x)]$$ where$$H_0(t)=\int_0^th_0(u)du$$ Thus, the distribution function is $$F(t|x)=1-exp[-H_0(t)exp(\beta'x)]$$ Let $Y$ be a random variable with distribution function $F$, then $U=F(Y)\sim U(0,1),\,(1-U)\sim U(0,1)$, i.e. $$U=exp[-H_0(t)exp(\beta'x)]\sim U(0,1)$$ if $h_0(t)>0$ for all $t$, then $H_0$ can be inverted and the survival time $T$ of the model can be written as $$T=H_0^{-1}[-log(U)exp(-\beta'x)]$$ where $U\sim U(0,1)$.

To simply the problem, here we only consider one covariate $x$, which indicates whether the sample belongs to the control arm ($x=0$) or the treatment arm ($x=1$), and set a negative $\beta$ under the assumption that the treatment has a consistent positive effect.

Now, we only need to know $H_0^{-1}$ to simulate the survival time. To do so, we consider two commonly used survival time distributions: **Exponential distribution** and **Weibull distribution**.

### Exponential Distribution

For exponential distribution with scale parameter $\lambda>0$, the possibility density function is $f_0=\lambda exp(-\lambda t)$.

Thus, $T=-\lambda^{-1}log(U)exp(-\beta'x)$ where $U\sim U(0,1)$.

### Weibull Distribution

For Weibull distribution with the scale parameter $\lambda$, and is the shape parameter $\gamma$, the possibility density function is $f_0=\lambda\gamma t^{\gamma-1}exp(-\lambda t^\gamma)$. Thus, $T={(-\lambda^{-1}log(U)exp(-\beta'x))}^{1/\gamma}$ where $U\sim U(0,1)$.

## Non-Proportional-Hazard Assumption

Under Non-Proportional-Hazard Assumption, we still consider the exponential model and Weibull model.

### Piecewise Exponential Model

**Late Effect:** We suppose the hazard function for the treatment arm is: $h(t|x=1)=\left\{ \begin{aligned} \lambda_0 & & t<1 \\ \lambda_1 & & t\geq 1 \end{aligned} \right.$

Thus, $T=\left\{ \begin{aligned} -\lambda_0^{-1}log(U) & & U>exp(-\lambda_0) \\ \frac{\lambda_1-log(U)}{\lambda_0+\lambda_1} & & U\leq exp(-\lambda_0) \end{aligned} \right.$

**Early Effect:** We can use the similar simulation method to generate piecewise exponential models in which the treatment arm shows early effect. The hazard function becomes: $h(t|x=1)=\left\{ \begin{aligned} \lambda_0 & & t\geq 1 \\ \lambda_1 & & t<1 \end{aligned} \right.$

Thus, $T=\left\{ \begin{aligned} -\lambda_0^{-1}log(U) & & U\leq exp(-\lambda_0) \\ \frac{\lambda_1-log(U)}{\lambda_0+\lambda_1} & & U>exp(-\lambda_0) \end{aligned} \right.$ where $U\sim U(0,1)$.

### Weibull Model

To simplify the problem, we assume the control and treatment arm share the same scale parameter $\lambda$. For the control arm, suppose the hazard function is: $h(t|x=0)=\lambda\gamma_0 t^{(\gamma_0-1)}.$ Thus, $T={(-\lambda^{-1}log(U))}^{1/\gamma_0}$.

Similarly, we can write the hazard function for the treatment arm as: $h(t|x=1)=\lambda\gamma_1 t^{(\gamma_1-1)}.$ We can derive that $T={(-\lambda^{-1}log(U))}^{1/\gamma_1}$.

# Simulation Results

## Proportional-Hazard Assumption

In this scenario, two tables have been presented, illustrating the baseline function following either an exponential **(Table 1)** or Weibull distribution **(Table 2)**. The results reveals that for assessing the Proportional-Hazard Assumption, the overall log-rank test (test 1_specificity) exhibits superior specificity, indicating more robust results. Furthermore, an increase in sample size tends towards higher specificity rates.

```{=tex}
\begin{center}
    \includegraphics[width=0.7\textwidth]{./graph/t1.png}
\end{center}
```
```{=tex}
\begin{center}
    \includegraphics[width=0.7\textwidth]{./graph/t2.png}
\end{center}
```
## Non-Proportional-Hazard Assumption

In the first scenario of addressing the Non-Proportional-Hazard Assumption, we assume a stepwise exponential distribution. The subsequent two tables demonstrate that employing the corresponding weighted log-rank test for late and early effects does indeed yield higher specificity rates. The weighted log-rank test (test 2_specificity) in **Table 3** demonstrates superior performance with increasing hazard ratios between the control and treatment groups for late effects. Likewise, the weighted log-rank test for early effects (test 3_specificity) in **Table 4** exhibits better efficacy with decreasing hazard ratios between the control and treatment groups. In addition, similar to the previous table, there is a tendency toward higher specificity rates as the sample size increases.

```{=tex}
\begin{center}
    \includegraphics[width=0.7\textwidth]{./graph/t3.png}
\end{center}
```
```{=tex}
\begin{center}
    \includegraphics[width=0.7\textwidth]{./graph/t4.png}
\end{center}
```

In the second scenario of addressing the Non-Proportional-Hazard Assumption, we assume that the treatment group and control group have different shape parameters $\gamma$ **(Table 5)**. In this case, the weighted log-rank test for late effects demonstrates the best performance, consistent with our intuitive findings. Furthermore, increasing the sample size continues to enhance specificity.

```{=tex}
\begin{center}
    \includegraphics[width=0.7\textwidth]{./graph/t5.png}
\end{center}
```
# Conclusion

In conclusion, our study sheds light on the performance of three hypothesis tests for time-to-event data analysis: the conventional log-rank test and two variations of the weighted log-rank test. We explored scenarios encompassing both proportional and non-proportional hazard functions, employing Monte Carlo simulation techniques to evaluate their power across various coefficient values.

Under the Proportional-Hazard Assumption, we found that the overall log-rank test consistently demonstrated superior specificity compared to the weighted log-rank tests.

When addressing the Non-Proportional-Hazard Assumption, our findings revealed nuances. In the first scenario with a stepwise exponential distribution, employing corresponding weighted log-rank tests for late and early effects resulted in higher specificity rates, highlighting the importance of selecting appropriate statistical methodologies. In the second scenario, where differing shape parameters $\gamma$ were assumed for treatment and control groups, the late weighted log-rank test emerged as the most effective, in line with expectations.

In addition, the trend of increasing specificity with larger sample sizes persisted, emphasizing the critical role of adequate sample sizes in ensuring reliable statistical inference.

Overall, our study offers valuable insights into choosing suitable statistical methodologies for time-to-event data analysis in clinical trials, especially under non-proportional hazard assumptions. These insights contribute to the continuous improvement of statistical practices in clinical research, ultimately enhancing the reliability and interpretability of trial results.

<!-- ## Graph -->

<!-- ### Proportional-Hazard Assumption -->

<!-- We gave examples to see the proportional hazard ratio functions by setting a exponential distributed hazard function under $\lambda=0.5,\,\beta=-0.5$ and a Weibull distributed hazard function under $\lambda=0.5,\,\gamma=1.2,\,\beta=-0.5$, shown as follows: -->

<!-- ```{r, echo = FALSE} -->

<!-- # Exponential Distribution -->

<!-- control_func_exp <- function(t) return(0.5) -->

<!-- treatment_func_exp <- function(t) return(0.5 * exp(-0.5)) -->

<!-- p1 <- hazard_plot(control_func_exp, treatment_func_exp, 5,  -->

<!--                   "Proportional-Hazard Function for Exponential Models") -->

<!-- # Weibull Distribution -->

<!-- control_func_weibull <- function(t) return(0.5 * 1.2 * t^(1.2 - 1)) -->

<!-- treatment_func_weibull <- function(t) return(0.5 * 1.2 * t^(1.2 - 1) * exp(-0.5)) -->

<!-- p2 <- hazard_plot(control_func_weibull, treatment_func_weibull, 5, -->

<!--                   "Proportional-Hazard Function for Weibull Models") -->

<!-- p1 <- p1 + ggtitle("Proportional-Hazard Function\nfor Exponential Models") -->

<!-- p2 <- p2 + ggtitle("Proportional-Hazard Function\nfor Weibull Models") -->

<!-- grid.arrange(p1, p2, nrow = 1) -->

<!-- ``` -->

<!-- ### Non-Proportional-Hazard Assumption -->

<!-- **Piecewise Exponential Model:** The hazard function under $\lambda_0=0.5,\,\lambda_1=0.3$ could reach late effect and hazard function under $\lambda_0=0.5,\,\lambda_1=0.3$ chould reach early effect, shown as follows: -->

<!-- ```{r, echo = FALSE} -->

<!-- control_func <- function(t) return(0.5) -->

<!-- treatment_func <- function(t) return(0.5 * (t < 1) + 0.3 * (t >= 1)) -->

<!-- p1 <- hazard_plot(control_func, treatment_func, 5, -->

<!--                   "Late-Effect Exponential Models") -->

<!-- control_func <- function(t) return(0.5) -->

<!-- treatment_func <- function(t) return(0.5 * (t >= 1) + 0.3 * (t < 1)) -->

<!-- p2 <- hazard_plot(control_func, treatment_func, 5, -->

<!--                   "Early-Effect Exponential Models") -->

<!-- grid.arrange(p1, p2, nrow = 1) -->

<!-- ``` -->

<!-- **Weibull Model:** The hazard function under $\lambda=0.5,\,\gamma_0=2,\,\gamma_1=1.5$ could be shown as follows: -->

<!-- ```{r, echo = FALSE} -->

<!-- control_func = function(t) return(0.5 * 2 * t ^ (2 - 1)) -->

<!-- treatment_func = function(t) return(0.5 * 1.5 * t ^ (1.5 - 1)) -->

<!-- hazard_plot(control_func, treatment_func, 5, -->

<!--             "Non-Proportional-Hazard Function for Late-Effect Weibull Models") -->

<!-- ``` -->

# Referrence

Austin, P. C. (2012). Generating survival times to simulate Cox proportional hazards models with time-varying covariates. Statistics in Medicine, 31(29), 3946--3958. <https://doi.org/10.1002/sim.5452>

Bender, R., Augustin, T., & Blettner, M. (2005). Generating survival times to simulate Cox proportional hazards models. Statistics in Medicine, 24(11), 1713--1723. <https://doi.org/10.1002/sim.2059>

Bardo, M. F., Huber, C., Benda, N., Brugger, J., Fellinger, T., Vaidotas Galaune, Heinz, J., Heinzl, H., Hooker, A. C., Florian Klinglmüller, Franz König, Mathes, T., Mittlböck, M., Posch, M., Ristl, R., & Friede, T. (2023). Methods for non-proportional hazards in clinical trials: A systematic review. ArXiv (Cornell University). <https://doi.org/10.48550/arxiv.2306.16858>
